---
title: "Model Building"
author: "Evan Bowman"
date: "2022-11-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressMessages(library(tidyverse))
```

# HSI Model 
## Data Loading and Selection
```{r}
# Loading in Data
hsei <- read_csv("data/Heat_Sensitivity_Exposure_Index.csv", show_col_types = F)
hsei$P_LIMENG = as.double(hsei$P_LIMENG)

# Isolating Socio Demographic Varibles and HSI
hsi_model <- hsei %>%
  select(P_POC:HSI)
```

## Linear Model Fitting using best Subset Regression
```{r}
library(leaps)
```

```{r}
# Initial Model Fitting
regFull <- regsubsets(HSI~ ., hsi_model)
summary(regFull)


# Regression Summary Plots
par(mfrow=c(1,1))
plot(regFull, scale = "r2")
plot(regFull, scale = "adjr2")
plot(regFull, scale = "Cp")
plot(regFull, scale = "bic")

summary(regFull)$adjr2  
summary(regFull)$cp  
summary(regFull)$bic  

# Identifying best values
which.max(summary(regFull)$adjr2)
which.min(summary(regFull)$cp) 
which.min(summary(regFull)$bic) 
```

From this initial best subsets method, model 5 returns the highest r2 value, lowest CP value, and is very close to the lowest BIC value. 

## Forward Stepwise
```{r}
# Stepwise Fitting
regfit.fwd <- regsubsets(HSI ~ ., data = hsi_model, method = "forward")
summary(regfit.fwd)

# Summary Plots
par(mfrow=c(1,1))
plot(regfit.fwd, scale = "r2")
plot(regfit.fwd, scale = "adjr2")
plot(regfit.fwd, scale = "Cp")
plot(regfit.fwd, scale = "bic")

summary(regfit.fwd)$adjr2  
summary(regfit.fwd)$cp  
summary(regfit.fwd)$bic  

# Identifying best values
which.max(summary(regfit.fwd)$adjr2)
which.min(summary(regfit.fwd)$cp) 
which.min(summary(regfit.fwd)$bic) 
```

From this initial forwards stepwise method, model 5 returns the highest r2 value, lowest CP value, and is very close to the lowest BIC value. 

## Backwards Stepwise
```{r}
# Backwards Fitting
regfit.bwd <- regsubsets(HSI ~ ., data = hsi_model, method = "backward")
summary(regfit.bwd)

# Summary Plots
par(mfrow=c(1,1))
plot(regfit.bwd, scale = "r2")
plot(regfit.bwd, scale = "adjr2")
plot(regfit.bwd, scale = "Cp")
plot(regfit.bwd, scale = "bic")

summary(regfit.bwd)$adjr2  
summary(regfit.bwd)$cp  
summary(regfit.bwd)$bic  

# Identifying best values
which.max(summary(regfit.bwd)$adjr2)
which.min(summary(regfit.bwd)$cp) 
which.min(summary(regfit.bwd)$bic) 
```

From this initial backwards stepwise method, model 5 returns the highest r2 value, lowest CP value, and is very close to the lowest BIC value. 

## Cross Validation of Models
```{r}
# Creating Validation Set
set.seed(1)
train <- sample(c(TRUE, FALSE), nrow(hsi_model),replace = TRUE)
val <- (!train)
```

```{r}
# Applying Best Subset Models to Training Data
regfit.best <- regsubsets(HSI ~ ., data = hsi_model[train, ])
```

# Compute the validation set error for the best model of each model size.
```{r}
# Compute the validation set error for the best model of each model size.
val.mat <- model.matrix(HSI ~ ., data = hsi_model[val, ], nvmax = 9) # This a model matrix from the validation data.
head(val.mat)
val.errors <- rep(NA, 9) # there are 9 variables and we will look at a model of each size (i.e num vars)

for (i in 1:9) {
 coefi <- coef(regfit.best, id = i) # get coef from the best model of size i
 pred <- val.mat[, names(coefi)] %*% coefi # predicted values = Xt*beta 
 val.errors[i] <- mean((hsi_model$HSI[val] - pred)^2) # MSE (validation observed y - yhat)^2
}

data.frame(val.errors,numvars=1:9)
plot(1:9,val.errors,xlab='num vars',ylab='CV errors',main='CV errors vs. num vars',pch=19,type='b')
which.min(val.errors)
coef(regfit.best,id=which.min(val.errors))
```

# Finally, we perform best subset selection on the full data set, and select the best seven-variable model.
# We use the full data to obtain more accurate coefficient estimates.  Note that we perform best subsets
# selection on the full data set and select the best seven-variable model, rather than simply using
# the variables that were obtained from the training set. This is because the best seven-variable model
# on the full data may (and here it does) differe from the corresponding model on the training data.

regfit.best <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19)
coef(regfit.best,7)
```

